---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 2,
  dpi = 150
)
library(ezTrack)
```

## Introduction

This vignette demonstrates a full workflow using ezTrack. It imports tracking data with non-standard column names, cleans, standardizes, and (optionally) converts it to a Spatial Feature, computes a summary table of the tracking data, and plots the tracks and home ranges on an interactive map.

## 1. Simulate Tracking Data with Irregular Format

We simulate a small tracking dataset and pass the dataset to `ez_track()`. If column names aren't provided, `ez_track()` will automatically attempt to guess which columns correspond to ID, timestamp, longitude, and latitude.

If automatic column guessing fails or guesses incorrectly, you can manually specify which columns represent the ID, timestamp, longitude, and latitude.

```{r}
set.seed(123)

n_individuals <- 4
n_points <- 20

sample_data <- data.frame(
  animal = rep(paste0("Animal_", LETTERS[1:n_individuals]), each = n_points),
  date = rep(seq.POSIXt(
    from = as.POSIXct("2023-01-01 08:00"),
    by = "36 hours",
    length.out = n_points), n_individuals),
  location_long = runif(n_individuals * n_points, 5.5, 7),
  location_lat = runif(n_individuals * n_points, 52, 53.5)
)

head(sample_data)

tracks <- ez_track(
  data = sample_data,
  as_sf = TRUE
)

tracks_manual <- ez_track(
  data = sample_data,
  as_sf = TRUE,
  id_col = 'animal',
  time_col = 'date',
  x_col = 'location_long',
  y_col = 'location_lat',
)
```

## 2. Summarize the Tracking Data

Summary statistics can be computed for each individual using `ez_summary()`. For easy use in reports or presentations, HTML reports can be generated using `report = TRUE`. Summaries can also be filtered by date using the `startDate` and `endDate` arguments.

The following summary statistics are returned for each unique id:

-   n_fixes: Number of location records\
-   first_location: Timestamp of the first recorded location\
-   last_location: Timestamp of the last recorded location\
-   tracking_duration_days: Duration between first and last fix (in days)\
-   fixes_per_day: Average number of fixes per day\
-   median_interval_hours: Median interval between fixes (in hours)\
-   max_time_gap_days: Longest time gap between consecutive fixes (in days)\
-   distance_km: Total distance traveled (in kilometers), calculated using the Haversine formula\
-   avg_speed_kmh: Average speed (km/h), computed as distance divided by tracking duration in hours

```{r}
summary_table <- ez_summary(tracks)
summary_table
```

## 3. Visualize the Tracks on an Interactive Map

We can display the movement tracks using an interactive Leaflet map with `ez_map()` with a simple command.

```{r}
ez_map(tracks)
```

Or with additional visualization arguments.

```{r}
ez_map(tracks, plotPath = TRUE,
  plotPoints = TRUE,
  pointRadius = 2,
  pointColor = "id",
  pointOpacity = 1,
  pointStroke = TRUE,
  pointStrokeColor = "black",
  pathColor = "id",
  pathWidth = 2,
  pathOpacity = 0.5,
  showLabels = TRUE)

```

As well as with date range filters.

```{r}
ez_map(tracks,
  startDate = "2023-01-01",
  endDate = "2023-01-10")

```

## 4. Estimate Home Ranges

Individual or population-level Minimum Convex Polygons (MCP) or Kernel Density Estimate (KDE) home ranges can be created with `ez_home_range()`.

```{r}
home_ranges_mcp <- ez_home_range(tracks, method = 'mcp', level = 50)
home_ranges_kde <- ez_home_range(tracks, method = 'kde', level = 95)

```

The result is an `sf` object containing one home range polygon per tracked individual or population (if `population = TRUE`).

Home ranges can be vizualised with `ez_map`:

```{r}
ez_map(home_ranges_mcp)
ez_map(home_ranges_kde, homeRangeColor = 'id')
```
## 5. Visualizing Latitude Over Time

`ez_latitude_plot()` can be used to visualize latitudinal movement (north-south) over time for each tracked animal. The plot supports:

- Optional faceting by individual

- Date axis customization

- Date range filtering

```{r}
ez_latitude_plot(tracks)

ez_latitude_plot(
  tracks,
  facet = TRUE,
  startDate = "2023-01-01",
  endDate = "2023-01-05",
  date_breaks = "1 day",
  date_format = "%d-%b",
)


```

## Developer Notes

To update vignette pkgdown site:

devtools::build_vignettes()
pkgdown::build_articles()
pkgdown::build_site()

