---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 2,
  dpi = 150
)
library(ezTrack)
```

## Introduction

This vignette demonstrates a full ezTrack workflow: importing tracking data with non-standard column names, cleaning and standardizing it, converting it to a spatial object, generating summary tables, visualizing fix rates and latitude over time, and plotting tracks and home ranges on an interactive map.

## 1. Import and format tracking data

We pass a dataset of GPS-tagged migratory shorebirds to `ez_track()`. If column names aren't provided, `ez_track()` will automatically attempt to guess which columns correspond to ID, timestamp, longitude, and latitude.


```{r}
data(godwit_tracks)
head(godwit_tracks)

tracks <- ez_track(
  data = godwit_tracks
)
```

If automatic column guessing fails or guesses incorrectly, you can manually specify which columns represent the ID, timestamp, longitude, and latitude.

```{r}
tracks_manual <- ez_track(
  data = godwit_tracks,
  id_col = 'individual.local.identifier',
  time_col = 'timestamp',
  x_col = 'location.long',
  y_col = 'location.lat',
)
```

You can also subsample the dataset using the `subsample()` argument, which limits the number of location fixes per time unit. Use a simple string format such as "1 per hour" or "2 per day" to control the sampling frequency.

```{r}
tracks_subsampled <- ez_track(
  data = godwit_tracks,
  subsample = "2 per hour"
)
```

## 2. Tracking data summary reports

Summary reports can be computed for each individual using `ez_summary()`. For easy use in reports or presentations, HTML reports can be generated using `report = TRUE`. Summaries can also be filtered by date using the `start_date` and `end_date` arguments.

The following summary statistics are returned for each unique id:

-   n_fixes: Number of location records\
-   first_location: Timestamp of the first recorded location\
-   last_location: Timestamp of the last recorded location\
-   tracking_duration_days: Duration between first and last fix (in days)\
-   fixes_per_day: Average number of fixes per day\
-   median_interval_hours: Median interval between fixes (in hours)\
-   max_time_gap_days: Longest time gap between consecutive fixes (in days)\
-   distance_km: Total distance traveled (in kilometers), calculated using the Haversine formula\
-   avg_speed_kmh: Average speed (km/h), computed as distance divided by tracking duration in hours

```{r}
summary_table <- ez_summary(tracks)
summary_table
```

## 3. Visualize latitude over time

`ez_latitude_plot()` can be used to visualize latitudinal movement (north-south) over time for each tracked animal. 


```{r}
ez_latitude_plot(tracks)
```

The plot supports:

- Optional faceting by individual

- Date axis customization

- Date range filtering

- `ggplot2` parameters for further customization.


```{r}
ez_latitude_plot(
  tracks,
  facet = TRUE,
  start_date = "2025-01-01",
  end_date = "2025-04-28",
  date_breaks = "2 months",
  date_format = "%b"
)
```

## 4. Visualize fix rate over time

`ez_fix_rate_plot()` can be used to visualize fix rates over time for each tracked animal, using a tick mark to represent each location fix. 


```{r}
ez_fix_rate_plot(tracks)
```

The plot supports:

- Date axis customization

- Date range filtering

- `ggplot2` parameters for further customization.


## 5. Map tracks and home ranges

We can display tracks using an interactive Leaflet map with `ez_map()` with a simple command.

```{r}
ez_map(tracks)
```

As well as with date range filters and a variety of style arguments.

```{r}
ez_map(tracks,
  start_date = "2025-04-10",
  end_date = "2025-04-28",
  point_color = "timestamp",
  point_size = 2,
  path_opacity = 0.3)

```

## 6. Estimate home ranges

Individual or population-level Minimum Convex Polygons (MCP) or Kernel Density Estimate (KDE) home ranges can be created with `ez_home_range()`.

```{r}
home_ranges_mcp <- ez_home_range(tracks, method = 'mcp', level = 95, start_date = '2025-04-12')

```

The result is an `sf` object containing one home range polygon per tracked individual or population (if `population = TRUE`).

Home ranges can be visualized with `ez_map`.

```{r}
ez_map(home_ranges = home_ranges_mcp)
```

Home ranges can also be mapped alongside tracking data. In this example, we use the start_date argument to ensure the tracking data aligns temporally with the home range period.

```{r}
ez_map(tracks, home_ranges = home_ranges_mcp, start_date = '2025-04-12')
```

## Developer notes
To update vignette pkgdown site:
pkgdown::build_site()

